---
title: 'Bios 6301: HW 2 Jordan Clark'
author: 'Jordan Clark'
output: pdf_document
---

*(informally) Due Thursday, 17 September, 1:00 PM*

50 points total.

This assignment won't be submitted until we've covered Rmarkdown.
Create R chunks for each question and insert your R code appropriately.
Check your output by using the `Knit PDF` button in RStudio.

1. **Working with data** In the `datasets` folder on the course GitHub repo, you will find a file called `cancer.csv`, which is a dataset in comma-separated values (csv) format. This is a large cancer incidence dataset that summarizes the incidence of different cancers for various subgroups. (18 points)

    1. Load the data set into R and make it a data frame called `cancer.df`. (2 points)
    ```{r}
    setwd("~/Desktop/Bios6301")
    getwd()
    cancer.df<-read.csv("./datasets/cancer.csv")
    head(cancer.df)
    names(cancer.df)
    ```
    2. Determine the number of rows and columns in the data frame. (2)
```{r}
   dim(cancer.df)
   #8 columns and 42120 rows
```

    3. Extract the names of the columns in `cancer.df`. (2)
```{r}
col.names<-c(colnames(cancer.df))
col.names    
```

    4. Report the value of the 3000th row in column 6. (2)
```{r}
print(cancer.df[3000,6])

```
    5. Report the contents of the 172nd row. (2)
```{r}
print(cancer.df[172,])

```
    6. Create a new column that is the incidence *rate* (per 100,000) for each row.(3)
```{r}
cancer.df<-transform(cancer.df, i.rate=incidence/100000 )
head(cancer.df)
```
    7. How many subgroups (rows) have a zero incidence rate? (2)
```{r}
zero<-0.0
zero    
cancer0.df<-cancer.df[cancer.df$i.rate %in% zero,]    
head(cancer0.df)
```
    8. Find the subgroup with the highest incidence rate.(3)
```{r}
which.max(cancer.df[,'i.rate'])
head(order(cancer.df[,'i.rate'],decreasing=T))
#Looks like row 21387
    cancer.df[21387,]
```
2. **Data types** (10 points)

    1. Create the following vector: `x <- c("5","12","7")`. Which of the following commands will produce an error message? For each command, Either explain why they should be errors, or explain the non-erroneous result. (4 points)

```{r}
x <- c("5","12","7")
class(x)
max(x)
#It returns a 7 because the elements are characters. 12 is not the max...
#... because it starts with 1. So max returns a pseudo-alphabetic order. 
sort(x)
#Because our elements are characters, '1' is like 'a', '5' is like 'e' ...
#... and '7' is like 'g', and they are ordered pseudo-alphabetically.
#(sum(x)) #I'm commenting this out because it won't compile otherwise.
#It is an error because you cannot sum characters like you can numerically.    
```
          
    2. For the next two commands, either explain their results, or why they should produce errors. (3 points)

```{r}
            y <- c("5",7,12)
    class(y[2])
 #           y[2] + y[3]
      #It appears that the elements of this vector have taken the class... 
    #...of the first element, which is a character. 
```

    3. For the next two commands, either explain their results, or why they should produce errors. (3 points)
```{r}
            z <- data.frame(z1="5",z2=7,z3=12)
            z[1,2] + z[1,3]
    #It appears that this worked because it's a data frame, rather than a vector.
```
3. **Data structures** Give R expressions that return the following matrices and vectors (*i.e.* do not construct them manually). (3 points each, 12 total)

    1. $(1,2,3,4,5,6,7,8,7,6,5,4,3,2,1)$
    ```{r}
    xx<-(1:8)
    yy<-rev(xx)
    zz<-append(xx,yy)
    zz<-zz[-8]
    zz
  #somewhat manually constructed?
```
  
    2. $(1,2,2,3,3,3,4,4,4,4,5,5,5,5,5)$
```{r}
rep(1:5, 1:5)
```

    3. $\begin{pmatrix}
      0 & 1 & 1  \\
      1 & 0 & 1  \\
      1 & 1 & 0  \\
    \end{pmatrix}$
```{r}  
  g<-c(0,1)
  f<-c(0,1)
  h<-c(1,0)
  i<-expand.grid(g,f,h)
  i
  #This is obviously too much. 
  i.i<-i[which(rowSums(i)==2),]
  i.i
    #Just reorder this puppy
  i.ii<-i.i[,c(2,1,3)]
    i.ii
    #Meandering solution but it worked. 
```
    4. $\begin{pmatrix}
      1 & 2 & 3 & 4 \\
      1 & 4 & 9 & 16 \\
      1 & 8 & 27 & 64  \\
      1 & 16 & 81 & 256 \\
      1 & 32 & 243 & 1024  \\
    \end{pmatrix}$
```{r}
n=5
g<-matrix(0,nrow=n,ncol=4)
a<-rep(1:n,1)
a
x<-c(1,2,3,4)
for (j in 1:n)
{g[j,]<-(x^j)
}
g
#Absolutely ridiculous. I'm shocked that I actually did that. 
```


4. **Basic programming** (10 points)

    1. Let $h(x,n)=1+x+x^2+\ldots+x^n = \sum_{i=0}^n x^i$. Write an R program to calculate $h(x,n)$ using a `for` loop. (5 points)
```{r}
x<-2
h<-2
n<-5
x.fun<-rep(0,n)
for(i in 0:(n-1)) {
  x.fun[i]<-x^(i+1)
  y<-1+x+sum(x.fun)
}
x.fun
y

1+h+h**2+h**3+h**4+h**5

sum(x^(0:n))
#Here's an easy non-loop way to do it. 
rowSums(outer(x, 0:n, "^"))
#Here's another super cool way to do it.
```
    1. If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Write an R program to perform the following calculations. (5 points)


1. Find the sum of all the multiples of 3 or 5 below 1,000. (3, [euler1])
```{r}        
#Here's my best shot
x<-numeric(999) 
#creating a vector that holds the values
for(i in 1:999) {
if (i %%3 == 0 || i%%5==0) {
x[i]<-i} else {x[i]<-0}}
x 
#this is a vector with all the multiples, with 0's in non-multiples
y<-x[x !=0] 
#remove the zero's
y 
#this is a vector with just the multiples
sum(y)
#The answer is 233168
```
1. Find the sum of all the multiples of 4 or 7 below 1,000,000. (2)
```{r}
#Same deal as before, just different multiples. 
a<-numeric(999999)
for(i in 1:999999){
if (i %%4 ==0 || i%%7==0){
a[i]<-i} else {a[i]<-0}
}
head(a)
b<-a[a !=0] 
head(b)
sum(b)
#answer is 178571071431
#You probably won't like the manual entry and deletion of 0 terms...
#... but this is the way I found that works. 
```
1. Each new term in the Fibonacci sequence is generated by adding the previous two terms. By starting with 1 and 2, the first 10 terms will be $(1, 2, 3, 5, 8, 13, 21, 34, 55, 89)$. Write an R program to calculate the sum of the first 15 even-valued terms. (5 bonus points, [euler2])
```{r}
#Going to try to determine the 15th even value term first
bin <- 45
fibvals <- numeric(bin)
fibvals[1] <- 1
fibvals[2] <- 1
for (i in 3:bin) { 
   fibvals[i] <- fibvals[i-1]+fibvals[i-2]
} 
fibvals
#Here are the first 45 values of the fibonacci sequence
#(I specified bin with some prior knowledge that fib. sequence produces...
#...an even number every 3rd term. Might be cheating a little)

c<-numeric(15)
for(i in seq(fibvals)) {
if(fibvals[i]%%2==0) { 
c[i]<-fibvals[i]} else{c[i]<-0}
}
head(c)
c
ci<-c[c !=0] 
length(ci)
ci
sum(ci)
#The answer is 1485607536. 
#Definitely not the most efficient way of doing it, but given that ...
#... I have almost no prior experience with loops, I'm happy with it. 
```
Some problems taken or inspired by projecteuler.

[euler1]:https://projecteuler.net/problem=1
[euler2]: https://projecteuler.net/problem=2


### Notes/Grade
```{r}
    
# Correction to Number 1.6
cancer.df[,'rate'] <- cancer.df[,'incidence']/cancer.df[,'population']*100000   

# Another way to do 3.1
c(seq(1,8), seq(7,1))

# Another way to do 3.3
+!diag(3)

# Another way to do 3.4
t(outer(1:4, 1:5, "^"))

# Another way to do 4.1
x <- 5
n <- 2
h <- 0
for(i in seq(0,n)) {
h <- h + x^i
}
h
    
# 4.2
sum(unique(c(seq(3,999,by=3), seq(5,999,by=5))))
sum(unique(c(seq(4,999999,by=4), seq(7,999999,by=7))))

# 4.3
r <- numeric(15)
r[1] <- 2
i <- 2
x1 <- 1
x2 <- 2
while(i <= 15) {
n <- x1 + x2
x1 <- x2
x2 <- n
  if(n %% 2 == 0) {
  r[i] <- n
  i <- i+1
  }
}
r 
sum(r)    
    
```

Grade:
48 points + 5 bonus points = 53/50 points